---
title: 'Statistical Sleuth in R: Chapter 2'
author: "Adam Loy"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document:
    number_sections: true
---

```{r setup, include=FALSE}
library(dplyr)
library(ggformula)
library(Sleuth3)
```


# Introduction

This document is intended to help describe how to undertake analyses introduced as examples in the Third Edition of the *Statistical Sleuth* (2013) by Fred Ramsey and Dan Schafer. More information about the book can be found at http://www.proaxis.com/~panorama/home.htm.  

This work adapts work done by Linda Loi, Ruobing Zhang, Kate Aloisio, and Nicholas J. Horton. Their work leveraged initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. 

In this chapter we need to load the following packages

```{r message=FALSE}
library(ggformula) # graphics
library(dplyr)     # data manipulation
library(Sleuth3)   # Sleuth data sets
```


# Evidence Supporting Darwin’s Theory of Natural Selection

Do birds evolve to adapt to their environments? That’s the question being addressed by Case
Study 2.1 in the *Sleuth*.

## Statistical summary and graphical display

```{r}
summary(case0201)
```

```{r}
case0201 %>%
  group_by(Year) %>%
  summarize(min = min(Depth), Q1 = quantile(Depth, probs = .25),
            median = median(Depth), Q3 = quantile(Depth, probs = .75),
            max = max(Depth), mean = mean(Depth), sd = sd(Depth))
```

A total of 178 subjects are included in the data: 89 are finches that were caught in 1976 and
89 are finches that were caught in 1978. The following are alternatives to Display 2.1 on page 30.

```{r fig.height=3, fig.width=5}
gf_boxplot(Depth ~ factor(Year), data = case0201) + 
  xlab("Year") +
  coord_flip()
```

```{r fig.height=3, fig.width=5}
gf_histogram(~Depth, data = case0201, bins = 15) +
  facet_wrap(~Year)
```

```{r fig.height=3, fig.width=5}
gf_density(~Depth, fill= ~factor(Year), data = case0201) +
  scale_fill_brewer("Year", palette = "Set2")
```

The distributions are approximately normally distributed, with some evidence for a long left tail.


## Inferential procedures (two-sample t-test)

We can get the results of “Summary of Statistical Findings” (page 29) by using the following code:

```{r}
 t.test(Depth ~ Year, var.equal = TRUE, data = case0201)
```


**Note:** The assumption that the variances are equal is a relatively strong one, so if there is any doubt, assume unequal variances.


# Anatomical Abnormalities Associated with Schizophrenia

Is the area of brain related to the development of schizophrenia? That’s the question being addressed
by case study 2.2 in the Sleuth.

##  Statistical summary and graphical display

We begin by reading the data and summarizing the variables.

```{r}
summary(case0202)
```

A total of 15 subjects are included in the data. There are 15 pairs of twins; one of the twins has
schizophrenia, and the other does not. So there are 15 affected subjects and 15 unaffected subjects.
The difference in area of left hippocampus of these pairs of twins is:

## Inferential procedures (paired t-test)

We can get the results displayed on page 32 by conducting a paired t-test:

```{r}
t.test(case0202$Unaffected, case0202$Affected, paired = TRUE)
```

