---
title: 'Statistical Sleuth in R: Chapter 21'
author: "Adam Loy"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=.9in 
output:
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, warning=FALSE}
options(width = 90)
knitr::opts_chunk$set(echo = TRUE, comment = NULL, fig.height = 3, fig.width = 4)
library(dplyr)
library(ggformula)
library(Sleuth3)
library(broom)
```


# Introduction

This document is intended to help describe how to undertake analyses introduced as examples in the Third Edition of the *Statistical Sleuth* (2013) by Fred Ramsey and Dan Schafer. More information about the book can be found at http://www.proaxis.com/~panorama/home.htm.  

In this chapter we need to load the following packages (remember, you will need to install packages you have never used before if you are using your own computer).

```{r message=FALSE, warning=FALSE}
library(dplyr)     # data manipulation
library(ggformula) # graphics
library(Sleuth3)   # Sleuth data sets
```

We will also set some options to improve legibility of graphs and output.
```{r eval=TRUE}
# display four significant digits by default and no significance stars
options(digits = 4, show.signif.stars = FALSE) 
```


# Island Size and Bird Extinctions

## EDA

Calculating the proportions and empirical logits...

```{r}
case2101 <- case2101 %>%
  mutate(prop = Extinct / AtRisk, 
         logit = log((Extinct + 0.5) / (AtRisk - Extinct + 0.5)))
head(case2101)
```


```{r display 21.2}
gf_point(logit ~ Area, data = case2101) %>%
  gf_refine(coord_trans(x="log10")) %>%
  gf_labs(x = "Area (sq. km, log scale)", y = "Logit")
```

## Fitting a logistic regression model

```{r display 21.6}
island_glm1 <- glm(prop ~ log(Area), data = case2101,
                   family = binomial, weights = AtRisk)
summary(island_glm1)
```


## Examining residuals

```{r display 21.5}
island_resids <- case2101 %>%
  select(Island, prop) %>%
  mutate(raw.resid = resid(island_glm1, type = "response"),
         pearson.resid = resid(island_glm1, type = "pearson"),
         dev.resid = resid(island_glm1, type = "deviance"))
head(island_resids)
```

## Deviance goodness-of-fit test

```{r}
deviance(island_glm1)
df.residual(island_glm1)
1 - pchisq(12.06, df = 16)
```

Note: the deviance statistic can also be found are the "Residual deviance" in the summary of the fitted logistic regression model. (You should look at the summary above to verify this.)


## Inference about model coefficients

```{r display 21.7}
tidy(island_glm1)
```

```{r}
beta1 <- coef(island_glm1)[2]
beta1

se <- sqrt(vcov(island_glm1)[2,2])
se

beta1 + c(-1, 1) * qnorm(.975) * se
```

Drop-in-deviance test
```{r}
island_reduced <- update(island_glm1, . ~ . - log(Area))
anova(island_reduced, island_glm1, test = "Chisq")
```


# Moth Coloration and Natural Selection


## EDA

Calculating the proportions and empirical logits...

```{r}
case2102 <- case2102 %>%
  mutate(prop = Removed / Placed,
         logit = log((Removed + 0.5) / (Placed - Removed + 0.5)))
head(case2102)
```

```{r display 21.4}
gf_point(logit ~ Distance, data = case2102, color = ~Morph) %>%
  gf_line() %>%
  gf_refine(scale_color_brewer(palette = "Set2")) %>%
  gf_labs(x = "Distance (km) from Liverpool",
          y = "Logit")
```


