---
title: 'Statistical Sleuth in R: Chapter 10'
author: "Adam Loy"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=.9in 
output:
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, warning=FALSE}
options(width = 90)
knitr::opts_chunk$set(echo = TRUE, comment = NULL, fig.height = 3, fig.width = 4)
library(dplyr)
library(ggformula)
library(Sleuth3)
library(stargazer)
```


# Introduction

This document is intended to help describe how to undertake analyses introduced as examples in the Third Edition of the *Statistical Sleuth* (2013) by Fred Ramsey and Dan Schafer. More information about the book can be found at http://www.proaxis.com/~panorama/home.htm.  

This work adapts work done by Linda Loi, Ruobing Zhang, Kate Aloisio, and Nicholas J. Horton. Their work leveraged initiatives undertaken by Project MOSAIC (http://www.mosaic-web.org), an NSF-funded effort to improve the teaching of statistics, calculus, science and computing in the undergraduate curriculum. 

In this chapter we need to load the following packages (remember, you will need to install packages you have never used before if you are using your own computer).

```{r message=FALSE, warning=FALSE}
library(ggformula) # graphics
library(Sleuth3)   # Sleuth data sets
library(broom)     # extract pieces of lm output
library(gridExtra) # arrange multiple plots on a page
library(stargazer) # tables to display/compare models
```

We will also set some options to improve legibility of graphs and output.
```{r eval=TRUE}
options(digits=4) # display four significant digits by default
```


# Galileo’s data on the motion of falling bodies

Galileo investigated the relationship between height and horizontal distance. This is the question addressed in case study 10.1 in the Sleuth.

## Data coding, summary statistics and graphical display

We begin by reading the data and summarizing the variables.

```{r}
summary(case1001)
```

There were a total of 7 trials of Galileo’s experiment. For each trial, he recorded the initial height
and then measured the horizontal distance as shown in Display 10.1 (page 272). We can start to explore this relationship by creating a scatterplot of Galileo’s horizontal distances versus initial heights. The following graph is akin to Display 10.2 (page 273).

```{r}
gf_point(Distance ~ Height, data = case1001) %>%
  gf_labs(x = "Horizontal distance (punti)", y = "Initial height (punti)")
```

## Models

The first model created is a cubic model as interpreted on page 273 and summarized in
Display 10.13 (page 291).

```{r}
galileo_lm1 <- lm(Distance ~ Height + I(Height^2) + I(Height^3), data = case1001)
summary(galileo_lm1)
```

We next decrease the polynomial for Height by one degree to obtain a quadratic model as
interpreted on page 273 and summarized in Display 10.7 (page 281). This model is used for most
of the following results.

```{r}
galileo_lm2 <- lm(Distance ~ Height + I(Height^2), data = case1001)
summary(galileo_lm2)
```

The following figure replicates Display 10.2 (page 273), displaying the fitted quadratic model.

```{r}
gf_point(Distance ~ Height, data = case1001) %>%
  gf_lm(formula = y ~ x + I(x^2)) %>%
  gf_labs(x = "Horizontal distance (punti)", y = "Initial height (punti)")

```

To obtain the expected values of $\widehat{\mu}({\rm Distance}|{\rm Height} = 0)$ and $\widehat{\mu}({\rm Distance}|{\rm Height} = 250)$, we use the `predict()` command with the quadratic model as shown in Display 10.7 (page 281).

```{r}
predict(galileo_lm2, newdata = data.frame(Height = c(0, 250)), interval = "confidence")
```

We can also verify the above confidence interval calculations with the following code:

```{r}
# Obtain the fitted value and SE
predict(galileo_lm2, newdata = data.frame(Height = c(0, 250)), interval = "confidence", se.fit = TRUE)

# "by-hand" calculation
355.5 + c(-1, 1) * qt(0.975, df = 4) * 6.625 
```

To verify numbers on page 284, an interval for the predicted values , we used the following code:

```{r}
predict(galileo_lm2, newdata = data.frame(Height = c(0, 250)), interval = "predict")
```

Lastly, we produced an ANOVA for the quadratic model interpreted on page 288 (Display
10.11). Notice that to get the entries for the "Regression" row on Display 10.11, you need to add the entries from the first two rows. This is because R displays sequential sums of squares and degrees of freedom. (Caution: Only the last p-value is valid!)

```{r}
anova(galileo_lm2)
```

#  Echolocation in bats

How do bats make their way about in the dark? Echolocation requires a lot of energy. Does it
depend on mass and species? This is the question addressed in case study 10.2 in the *Sleuth*.

## Data coding, summary statistics and graphical display

We begin by reading the data, performing transformations where necessary and summarizing the
variables.

```{r}
case1002$Type <-factor(case1002$Type, 
                       levels = c("non-echolocating bats", "non-echolocating birds", 
                                  "echolocating bats"))
summary(case1002)
```

A total of 20 flying vertebrates were included in this study. There were 4 echolocating bats,
4 non-echolocating bats, and 12 non-echolocating birds. For each subject their mass and flight
energy expenditure were recorded as shown in Display 10.3 (page 274).

Next, we explore the relationship between log(energy expenditure) as a function of log(body
mass) for each group with a scatterplot. The following figure reproduces Display 10.4 (page 275).

```{r}
gf_point(log(Energy) ~ log(Mass), data = case1002, shape = ~Type) %>%
  gf_labs(x = "Body Mass (g) (log scale)", y = "Energy Expenditure (W) (log scale)") %>%
  gf_theme(legend.position=c(.2, .9)) # 0,0 is the bottom left; 1,1 is the top right
```

## Multiple regression

We first evaluate a multiple regression model for log(energy expenditure) given type of species and log(body mass) as defined on page 276 and shown in Display 10.6 (page 277).

```{r}
energy_lm1 <- lm(log(Energy) ~ log(Mass) + Type, data = case1002)
summary(energy_lm1)
```

Next, we calculate confidence intervals for the coefficients that are interpreted on page 278.

```{r}
confint(energy_lm1)
exp(confint(energy_lm1))
```

Since the significance of a model depends on which variables are included, *Sleuth* proposes
two other models, one only looking at the type of flying animal and the other allows the three
groups to have different straight-line regressions with mass. These two models are displayed below and discussed on pages 278-279.

```{r}
energy_lm2 <- lm(log(Energy) ~ Type, data = case1002)
energy_lm3 <- lm(log(Energy) ~ log(Mass) * Type, data = case1002)
```

Table 1 presents the models displays on page 279 using `stargazer`.

```{r, results='asis'}
stargazer(energy_lm2, energy_lm1, energy_lm3, type = "latex", digits = 2, 
          omit.table.layout = "sn", 
          title = "Coefficients and standard errors for the three models on pare 279 in Sleuth.", 
          header=FALSE, 
          covariate.labels = c("lmass", "bird", "ebat", "bird:lmass", "ebat:lmass"))
```


Next we can assess the model by evaluating the extra sums of squares F-test for testing the
equality of intercepts in the parallel regression lines model as shown in Display 10.10 (page 287).

```{r}
energy_slr <- lm(log(Energy) ~ log(Mass), data = case1002)
anova(energy_slr, energy_lm1)
```

We can also compare the full model with interaction terms and the reduced model (without
interaction terms) with the extra sum of squares F-test as described in Display 10.12 (page 290).

```{r}
anova(energy_lm1, energy_lm3)
```

## Linear combinations of regression coefficients

Finally, we can obtain the standard error for $\beta_2 - \beta_3$, as detailed on Display 10.15 (page 294), using the below commands:

```{r}
# 1. estimate the linear combination
estimate <- coef(energy_lm1)[3] - coef(energy_lm1)[4]
estimate

# 2. Obtain the variance-covariance matrix of the coefficients
V <- vcov(energy_lm1)
V

# 3. Calculate the estimated variance of the linear combination
v.b2 <- V[3, 3]
v.b2

v.b3 <- V[4, 4]
v.b3

cov.b2b3 <- V[3, 4]
cov.b2b3

estimated_var <- v.b2 + v.b3 - 2 * cov.b2b3
estimated_var

se <- sqrt(estimated_var)
se
```

